Survey of Cache Oblivious B-trees

### 2011-12-12 paper.tex
edits of streaming section + fix a lost citation
I lost a citation from merging- oops

In order for B-trees to remain versatile, they must be able to accommodate
various forms of data. However, traditional B-trees and string B-trees have
suboptimal performance when variable length strings are used as keys. In
addition, string B-trees do not support compression, and commonly used front
compression is handled sub-optimally. Regular B-trees and string B-trees are
not cache oblivious, and thus incur poorer performance as described
previously. Range queries have poor performance in B-trees. In
\cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
Oblivious String B-tree (COSB-tree), which they argue works efficiently in
these problem areas for variable length string keys. They also provide a
modified front-compression scheme, which gives their approach the advantages
of using less space and having improved locality within the compressed data.

\subsection{Description}
Bender et al. perform 2 experiments in \cite{BenderFaKu06}, in which a regular B-tree implementation tuned to various block sizes is compared to a static COSB-tree implementation. The experiment involved measuring the amount of time required to perform: 440,000 and 450,000 inserts of random values, a range query of all values, and 1000 random searches. The results of the experiment show that COSB-trees
outperformed B-trees most of the time for random searches, range queries and
random inserts. However, some point between 440,000 and 450,000 inserts the
COSB-tree had to rebalance the entire tree, causing it to fall behind in this
measurement. However, Bender et al. argue that the experiment was biased
towards the B-tree implementation, since the trees are "young", and thus still
well organized. The COSB-tree implementation is also compared to Berkeley DB
\cite{BerkeleyDB}, and achieves performance comparable to Berkeley DB with
tuned parameters. The results seem to show that there is a window in which
large and small-block size B-trees perform poorly, and that COSB-trees may be
a reasonable alternative to improve performance.

@@

The Dynamic COSB-tree consists of 3 major components: a centroid tree at the
top; a PMA structure of hash data for the keys known as
the \emph{hashdata} layer; and on the bottom a PMA structure for the original
key values, compressed using an augmented version of forward compression known as the \emph{keydata} layer. Each leaf of the centroid tree contains a pointer into the hashdata array. The centroid tree only contains a subset of the keys, so each leaf points to the respective \emph{neighborhood} of the key being searched for. After traversing the centroid tree and following the pointer into the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a sequential scan is performed in the hashdata array layer to find $k$. Each entry in the hashdata array contains the hash value of the corresponding key, and a pointer to the compressed key value in the keydata layer.

In the following subsections, centroid trees are explained, and a summary of
how to modify them to preserve balance and structure in the dynamic case is
provided. Two iterations on forward compression are also summarized to show
how to incrementally add support for variable length strings as keys for
dynamic COSB-trees.

@@

COSB-trees are a solution to the dictionary matching problem. Dictionary matching, or string matching, refers to matching strings using prefix search. Let $D$ be a dictionary of $N$ keys. To perform faster searches, the keys in $D$ are preprocessed using hashing and divide-and-conquer via centroid trees. Figure \ref{fig:centroid} depicts an example centroid tree.

Consider the compacted\footnote{where compacted means all non-branching paths are reduced to edges.} trie $T$ constructed over $D$ for string matching.
Consider the vertex $\rho$ in $T$ such that $\rho$ has
between N/3 and 2N/3 descendants, inclusive. Call $\rho$ the \emph{centroid vertex} of $T$. When searching for a key $k$ in
the compacted trie, you traverse the trie as follows: for each node $y$ that
you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on into the
trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the trie
resulting from excluding $y$ and its subtree. The \emph{centroid tree} is
obtained by rooting the tree at the centroid vertex $\rho$, and making its
children recursively defined centroid trees of its up and down tries.

The centroid tree in the COSB-tree is layed out in memory using a modified van Emde Boas (vEB) layout for
weight-balanced binary trees. The \emph{hyperhyperfloors} of nodes are used as
a guide for cutting the tree into equal-size subtrees for the vEB layout. The
\emph{hyperhyperfloor} of x is denoted as $\lhyperfloor{}x\rhyperfloor{}$, and
refers to rounding x down to the nearest power of a power of 2. This ensures
trees of equal size on all \emph{levels of detail}, as the centroid tree is
traversed.

The following approach is used to rebalance the centroid tree: Put all the
elements in DFS or Euler tour order for the compacted trie. Given this order, scan the trie to identify the centroid vertex. Then partition the trie into its respective
up and down tries, and recurse on the up and down tries. The Euler tour order
can be found in $O(\log{}N)$ scans, since the trie is only $O(\log{}N)$
deep. Thus these updates have amortized cost of $O(1+\log^{2}N/B)$ memory transfers.

The following approach is used for maintaining the vEB layout of the centroid
tree in the COSB-tree: first, weight balance is relaxed from a
maximum weight difference of 2 between left and right subtrees, to a larger
constant $c$. Thus a node $v$ only becomes out of balance every
$\Omega(\textsc{weight(v)})$ insertions or deletions of descendants of v. The nodes are stored in a PMA, preserving order of the nodes using $\Theta(N)$space.

On insert or delete, the area around the update point is scanned for sections of the array to rearrange. If a section around the update point reaches a certain "density" threshold, the nodes are evenly
distributed across the section to make room for more nodes. To
ensure that updating pointers isn't too expensive, any node included in the
section must have all of its subtrees included. The subtrees are dictated by the vEB layout, and subtrees are chosen from the bottom up. The smallest neighboring trees (single nodes) in the vEB layout are added first, and larger trees are included when all of their subtrees have been added to the section. Root trees are only added after all bottom trees are added (in terms of the vEB layout). Figure \ref{fig:centroid_relayout} gives an example of how subtrees are chosen for rearranging.

This approach
results in an amortized update cost of $O(1+(\log{}N)/B)$ memory transfers per
update. In conjunction with the cost of rebuilds described above, this is only
an additive $O(\log{}N/B)$ factor, so the overall bound is amortized
$O(1+\log^{2}N/B)$ memory transfers.
\begin{figure}

@@

\caption{A centroid layout example. Assume the smallest subtrees in this example only contain a single node. Suppose a node was inserted into a leaf position in the bottom left of the tree. The numbers represent the order in which each subtree is added to the section to be rearranged.}
\label{fig:centroid_relayout}
\end{figure}
\subsubsection{Modifying Front Compression to Preserve Locality}

This section provides a brief summary of how front compression is modified to
optimize compression and decoding for dynamic COSB-trees. Consider a sequence of $j$ keys to store $k_{1}$, $k_{2}$,
..., $k_{j}$. Without compression, this requires $\sum_{j}||k_{j}||$ space in
memory. To compress these keys using standard front compression, let $\pi_{j}$ be the
longest common prefix between $k_{j}$ and $k_{j+1}$. Now instead we store the
keys in the following format:

\begin{center}
$k_{1}$,$||\pi_{2}||$,$\sigma_{2}$,$||\pi_{3}||$$\sigma_{3}$,...,$||\pi_{j}||$,$\sigma_{j}$
\end{center}

where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$
bits. This saves almost $\sum_{j}||\pi_{j}||$ memory. However, since decoding
$k_{j}$ involves concatenating the first $\pi_{j}$ bits of $k_{j-1}$ to
$\sigma_{j}$, $k_{j-1}$ must be decoded. $k_{j-1}$ may depend on previous keys
as well, making $k_{j}$ expensive to decode.

Bender et al. provide a modification that reduces the high decoding cost of front compressing D to $O(1+||k||B)$, while maintaining a compressed size of $O(\llangle{D}\rrangle{})$\footnote{$\llangle{D}\rrangle{}$ is shorthand for the size of $D$ after front compression} after compression. The modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be
the first key added to the compression stream. For each subsequent key
$k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters,
add $\pi_{i}$,$\sigma_{i}$ as described above. Otherwise, add $k_{i}$ to the stream without compressing it. By construction, this scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.

@@

The operations supported by the COSB-tree are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
and \Range{}. The behavior of each operation for dynamic COSB-trees is
described below. \Insertkonly{} and \Delete{} are also supported but not
described here.

\Search{}: works as follows for a given key $k$. Compute the hash value for
$k$ and call it $\rho$, and traverse the centroid tree of the COSB-tree using
$\rho$ to obtain the successor $\rho_{s}$ and predecessor $\rho_{p}$ of $k$.
Index into the hashdata layer and perform a sequential scan across the PMA
structure from $\rho_{p}$ to $\rho_{s}$ to find $\rho$. On a match with value
$\rho'$, index into the keydata layer and decode $\rho'$ to obtain the
resulting key $k'$. If $k' = k$, return true. If $k' \ne k$ or $k$ was not
found during the sequential scan return false.

\Pred{},\Succ{}: works in the same way as \Search{} until the sequential scan
in the hashdata layer. When scanning the hashdata layer, look for $\rho_{p}$
or $\rho_{s}$, respectively, in place of $\rho$, and compare the appropriate
keys in the keydata layer.

\Range{}: This operation is not explicitly described in \cite{BenderFaKu06}, but presumably this is very similar to
\Search{}. Let $k_{b}$ be the beginning of the range and let $k_{e}$ be the
end of the range. Hash the range keys and identify the predecessor of
$\rho_{b}$ and the successor of $\rho_{e}$ using the centroid tree. Scan the
hashdata layer across from $\rho_{b,p}$ to $\rho_{e,s}$. Only start noting the
beginning of range when the first key with hash value after $\rho_{b}$, and
stop once we reach the first key with hash value after $\rho_{e}$. Decode and
return the set of keys within the given range in the keydata layer.

\subsection{Complexity}
The keys in are hashed and hash values are compared in place of the original key values to speed up comparisons in the centroid tree and hashdata layer. To avoid mismatches due to overlapping hash values, the original key values are compared only once at the bottom layer, resulting an additive $O(||k||/B)$ memory transfers in each of the operators for each key $k$ that needs to be compared.

\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high
probability (w.h.p.).  The centroid tree is stored in a PMA with a dynamic CO layout. This layout reduces the number of memory transfers required when traversing the tree to $O(log_{B}N)$.  Once in the hashdata layer, we are already within $O(logN)$ keys of $k$, which require $O(\log{}N)/B)$ memory transfers to scan. Decoding and comparing the original key values requires $O(1+||k||/B)$ memory transfers, due to the modified front compression scheme.

The \Pred{} and \Succ{} operations are similar in behavior to searches, and thus only have the added cost of having to operate on $k'$, where $k'$ is the predecessor or successor, respectively. These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers w.h.p.. Similarly, range queries take
$O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers,
where in this case $k'$ is the other end of the range and $Q$ is the set of
keys within the given range.

@@

@@

In addition, four non-blocking primitives are introduced:
\begin{inparaenum}[(a)]
  \item \textit{move},
  \item \textit{cell-insertion},
  \item \textit{cell-deletion}, and
  \item \textit{read}
\end{inparaenum}. These primitives are self-explanatory, except they
all work in a non-blocking manner (they can fail at any point, and it is
up to the current process to restart the operation on failure).

>>>>

In order for B-trees to remain versatile, they must be able to accommodate
various forms of data. However, traditional B-trees and string B-trees have
suboptimal performance when variable length strings are used as keys. In
addition, string B-trees do not support compression, and commonly used front
compression is handled sub-optimally. Regular B-trees and string B-trees are
not cache oblivious, and thus incur poorer performance as described previously.
In \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
Oblivious String B-tree (COSB-tree), which they argue works efficiently for
variable length string keys. They also provide a modified front-compression
scheme, which gives their approach the advantages of using less space and
having improved locality within the compressed data.

@@

The dynamic COSB-tree consists of three major components: a centroid tree at
the top, a PMA structure of hash data for the keys known as the \emph{hashdata}
layer; and on the bottom a PMA structure for the original key values,
compressed using an augmented version of forward compression known as the
\emph{keydata} layer. Each leaf of the centroid tree contains a pointer into
the hashdata array. The centroid tree only contains a subset of the keys, so
each leaf points to the respective \emph{neighborhood} of the key being
searched for. After traversing the centroid tree and following the pointer into
the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
sequential scan is performed in the hashdata array layer to find $k$. Each
entry in the hashdata array contains the hash value of the corresponding key,
and a pointer to the compressed key value in the keydata layer.

In the following sections, centroid trees are described, and a summary of
how to modify them to preserve balance and structure in the dynamic case is
provided. Two iterations on forward compression are also summarized to show
how to incrementally add support for variable length strings as keys for
dynamic COSB-trees.

@@

COSB-trees are a solution to the \textit{dictionary matching problem}.
Dictionary matching, or string matching, refers to matching strings using
prefix search. Let $D$ be a dictionary of $N$ keys. To perform faster searches,
the keys in $D$ are preprocessed using hashing and divide-and-conquer via
centroid trees. Figure \ref{fig:centroid} depicts an example centroid tree.

Consider the compacted\footnote{Compacted means all non-branching paths are
reduced to edges.} trie $T$ constructed over $D$ for string matching.  Consider
the vertex $\rho$ in $T$ such that $\rho$ has between N/3 and 2N/3 descendants,
inclusive. Call $\rho$ the \emph{centroid vertex} of $T$. When searching for a
key $k$ in the compacted trie, you traverse the trie as follows: for each node
$y$ that you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on
into the trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the
trie resulting from excluding $y$ and its subtree. The \emph{centroid tree} is
obtained by rooting the tree at the centroid vertex $\rho$, and making its
children recursively defined centroid trees of its up and down tries.

The centroid tree in the COSB-tree is layed out in memory using a modified van
Emde Boas (vEB) layout for weight-balanced binary trees. The
\emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into
equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x is
denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to
the nearest power of a power of 2. This ensures trees of equal size on all
\emph{levels of detail}, as the centroid tree is traversed.
% Should explain what "level of detail" means here

The following approach is used to rebalance the centroid tree: put all the
elements in DFS or Euler tour order for the compacted trie. Given this order,
scan the trie to identify the centroid vertex. Then partition the trie into its
respective up and down tries, and recurse on the up and down tries. The Euler
tour order can be found in $O(\log{N})$ scans, since the trie is only
$O(\log{N})$ deep. Thus these updates have amortized cost of
$O(1+(\log^{2}{N})/B)$ memory transfers.

The following approach is used for maintaining the vEB layout of the centroid
tree in the COSB-tree: first, weight balance is relaxed from a maximum weight
difference of 2 between left and right subtrees, to a larger constant $c$. Thus
a node $v$ only becomes out of balance every $\Omega(\textsc{weight(v)})$
insertions or deletions of descendants of $v$. The nodes are stored in a PMA,
preserving order of the nodes using $\Theta(N)$ space.

On insert or delete, the area around the update point is scanned for sections
of the array to rearrange. If a section around the update point reaches a
certain \textit{density} threshold, the nodes are evenly distributed across the
section to make room for more nodes. To ensure that updating pointers is not too
expensive, any node included in the section must have all of its subtrees
included. The subtrees are dictated by the vEB layout, and subtrees are chosen
from the bottom up. The smallest neighboring trees (single nodes) in the vEB
layout are added first, and larger trees are included when all of their
subtrees have been added to the section. Root trees are only added after all
bottom trees are added (in terms of the vEB layout). Figure
\ref{fig:centroid_relayout} gives an example of how subtrees are chosen for
rearranging.

This approach results in an amortized update cost of $O(1+(\log{N})/B)$ memory
transfers per update. In conjunction with the cost of rebuilds described above,
this is only an additive $O((\log{N})/B)$ factor, so the overall bound is
amortized $O(1+(\log^{2}N)/B)$ memory transfers.
\begin{figure}

@@

\caption{A centroid layout example. Assume the smallest subtrees in this
example only contain a single node. Suppose a node was inserted into a leaf
position in the bottom left of the tree. The numbers represent the order in
which each subtree is added to the section to be rearranged.}
\label{fig:centroid_relayout}
\end{figure}

\subsubsection{Modifying Front Compression to Preserve Locality}

This section provides a brief summary of how front compression is modified to
optimize compression and decoding for dynamic COSB-trees. Consider a sequence
of $j$ keys to store $k_{1},k_{2},...,k_{j}$. Without compression, this
requires $\sum_{j}||k_{j}||$ space in memory. To compress these keys using
standard front compression, let $\pi_{j}$ be the longest common prefix between
$k_{j}$ and $k_{j+1}$. Now instead we store the keys in the following format:
\begin{equation}
  k_{1}||\pi_{2}||\sigma_{2}||\pi_{3}||\sigma_{3}||...||\pi_{j}||\sigma_{j}
\end{equation} 
where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$
bits. This saves almost $\sum_{j}||\pi_{j}||$ memory. However, since decoding
$k_{j}$ involves concatenating the first $\pi_{j}$ bits of $k_{j-1}$ to
$\sigma_{j}$, $k_{j-1}$ must be decoded. $k_{j-1}$ may depend on previous keys
as well, making $k_{j}$ expensive to decode.

Bender et al. provide a modification that reduces the high decoding cost of
front compressing $D$ to $O(1+||k||B)$, while maintaining a compressed size of
$O(\llangle{D}\rrangle{})$\footnote{$\llangle{D}\rrangle{}$ is shorthand for
the size of $D$ after front compression} after compression. The modification
works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be the first key
added to the compression stream. For each subsequent key $k_{i}$, if $k_{i}$
can be decoded with just the last $c||k_{i}||$ characters, add
$\pi_{i}$, $\sigma_{i}$ as described above. Otherwise, add $k_{i}$ to the stream
without compressing it. By construction, this scheme guarantees that any key
$k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.

@@

The operations supported by the COSB-tree are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
and \Scan{}. The behavior of each operation for dynamic COSB-trees is
described below. \Insertkonly{} and \Delete{} are also supported but not
described here.

\Search{} works as follows for a given key $k$. Compute the hash value for
$k$ and call it $\rho$, and traverse the centroid tree of the COSB-tree using
$\rho$ to obtain the successor $\rho_{s}$ and predecessor $\rho_{p}$ of $k$.
Index into the hashdata layer and perform a sequential scan across the PMA
structure from $\rho_{p}$ to $\rho_{s}$ to find $\rho$. On a match with value
$\rho'$, index into the keydata layer and decode $\rho'$ to obtain the
resulting key $k'$. If $k' = k$, return true. If $k' \ne k$ or $k$ was not
found during the sequential scan return false.

\Pred{}, \Succ{} work in the same way as \Search{} until the sequential scan in
the hashdata layer. When scanning the hashdata layer, look for $\rho_{p}$ or
$\rho_{s}$, respectively, in place of $\rho$, and compare the appropriate keys
in the keydata layer.

\Scan{}: This operation is not explicitly described in \cite{BenderFaKu06}, but
presumably this is very similar to \Search{}. Let $k_{b}$ be the beginning of
the range and let $k_{e}$ be the end of the range. Hash the range keys and
identify the predecessor of $\rho_{b}$ and the successor of $\rho_{e}$ using
the centroid tree. Scan the hashdata layer across from $\rho_{b,p}$ to
$\rho_{e,s}$. Only start noting the beginning of range when the first key with
hash value after $\rho_{b}$, and stop once we reach the first key with hash
value after $\rho_{e}$. Decode and return the set of keys within the given
range in the keydata layer.

\subsection{Complexity}
The keys are hashed and hash values are compared in place of the original key
values to speed up comparisons in the centroid tree and hashdata layer. To
avoid mismatches due to hash collisions, the original key values are compared
only once at the bottom layer, resulting an additive $O(||k||/B)$ memory
transfers in each of the operators for each key $k$ that needs to be compared.

\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high probability
(w.h.p.).  The centroid tree is stored in a PMA with a dynamic CO layout. This
layout reduces the number of memory transfers required when traversing the tree
to $O(\log_{B}{N})$.  Once in the hashdata layer, we are already within $O(\log{N})$
keys of $k$, which require $O((\log{N})/B)$ memory transfers to scan. Decoding
and comparing the original key values requires $O(1+||k||/B)$ memory transfers,
due to the modified front compression scheme.

The \Pred{} and \Succ{} operations are similar in behavior to searches, and
thus only have the added cost of having to operate on $k'$, where $k'$ is the
predecessor or successor, respectively. These operations as a result take
$O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers (w.h.p.). Similarly, range
queries take $O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory
transfers, where in this case $k'$ is the other end of the range and $Q$ is the
set of keys within the given range.

@@

\subsection{Experimental Results}
Bender et al. perform two experiments in \cite{BenderFaKu06}, in which a
regular B-tree implementation tuned to various block sizes is compared to a
static COSB-tree implementation. The experiment involved measuring the amount
of time required to perform:
\begin{inparaenum}[(a)]
  \item 440,000 and 450,000 inserts of random values, 
  \item a range query of all values, and 
  \item 1000 random searches.
\end{inparaenum} The results of the
experiment show that COSB-trees outperformed B-trees most of the time for
random searches, range queries and random inserts. However, some point between
440,000 and 450,000 inserts the COSB-tree had to rebalance the entire tree,
causing it to fall behind in this measurement. However, Bender et al. argue
that the experiment was biased towards the B-tree implementation, since the
trees are ``young'', and thus still well organized. The COSB-tree
implementation is also compared to Berkeley DB \cite{BerkeleyDB}, and achieves
performance comparable to Berkeley DB with tuned parameters. The results seem
to show that there is a window in which large and small-block size B-trees
perform poorly, and that COSB-trees may be a reasonable alternative to improve
performance.

@@

In addition, four non-blocking primitives are introduced:
\begin{inparaenum}[(a)]
  \item \textit{move},
  \item \textit{cell-insertion},
  \item \textit{cell-deletion}, and
  \item \textit{read}.
\end{inparaenum} These primitives are self-explanatory, except they
all work in a non-blocking manner (they can fail at any point, and it is
up to the current process to restart the operation on failure).
